Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-17 paper/paper.tex
Referred to Thomas's paper and marked all our changes in blue

\added{While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).   }   \todelete{Designing  a  liveness  based  garbage
  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
Because  of   these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for a
eager language. An earlier  effort in efficiently garbage collecting
closures~\cite{Thomas19951}  used  a   similar  approach  where  every
closure  would maintain  a  sequence of  executable  code called  {\em
  preserver}  which  identified  exactly which  closure  variables  to
copy. The main focus of the  above work was also in garbage collecting
closures without using  an explicit stack. However, it not clear whether
their method could handle partially live data structures.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are:
\added{
\begin{itemize}
\item {\color{Myblue}We  propose a liveness  based GC scheme  for a lazy  first order
  functional language and  prove its correctness.  To the  best of our
  knowledge,  this is  the first  work on  liveness-based GC  for lazy
  languages.       All      previous      attempts~\cite{shaham01heap,
    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
    karkare06effectiveness} have  involved either imperative  or eager
  functional languages.  The novelty of our approach is in handling of
  liveness  of closures,  an  issue  that arises  due  to laziness  at
  runtime.}

@@

Note that this equation can also be viewed as a CFG with \{\acdr,
\clazy\} as terminal symbols and \Df{\length}{1} as the sole
non-terminal.

%\subsection{Generating liveness equations \Lv\  for function bodies}
\paragraph{Handling user-defined functions:}
\label{sec:bodylivenessbodies}
To  avoid analyzing  the body  of a  function for  each
call, we calculated the  liveness for the arguments and
the variables in a function  with respect to a symbolic
demand  $\sigma$.   To  get   the  actual  liveness  we
calculate an  over-approximation of the  actual demands
made by  all the  calls and  calculate the  liveness at
each  GC  point  inside  the  function  based  on  this
approximation.  The {\em summary demand} is
calculated by  taking a union  of the demands  at every
call site of a function analogous to 0-CFA~\cite{Shivers:1988}.

@@

\input{cfg-to-nfa}

  \SetStartEndCondition{ }{}{}%
  \SetKwProg{Pro}{procedure}{\string:}{}
  \SetKwProg{Fn}{function}{\string:}{}
  \SetKwFunction{Range}{range}%%
  \SetKw{KwTo}{in}\SetKwFor{For}{for}{\string:}{}%
  \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
  \SetKwFor{While}{while}{:}{fintq}%
  \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
  \SetKwFunction{Lgc}{$\mathsf{lgc}$}%
  \SetKwFunction{Copy}{$\mathsf{copy}$}%
  \SetKwFunction{LCopy}{$lgcCopy$}%
  \SetKwFunction{RCopy}{$rgcCopy$}%

>>>>

\added{While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).   }   \todelete{Designing  a  liveness  based  garbage
  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
Because  of   these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for a
eager language.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are:
\added{
\begin{itemize}
\item {\color{Myblue}We propose a liveness based GC scheme for a lazy
  first order functional language and prove its correctness.
  Thomas~\cite{Thomas19951} describes a copying garbage collector for
  Three Instruction Machine (TIM) that only traces live variables in a
  closure's environment, i.e., the variables that are useful for
  evaluating the closure.  Their notion of liveness of variables is
  close to the classical notion of liveness, and does not take into
  account liveness of parts of structures (say, only the spine of a
  list).  To the best of our knowledge, it is the only earlier work on
  GC that takes liveness for closures into account.  All previous
  attempts~\cite{shaham01heap, ran.shaham-sas03, shaham02estimating,
    asati14lgc, karkare06effectiveness} have involved either
  imperative or eager functional languages.  The novelty of our
  approach is in handling of liveness of closures, an issue that
  arises due to laziness at runtime.}

@@

Note that this equation can also be viewed as a CFG with \{\acdr,
\clazy\} as terminal symbols and \Df{\length}{1} as the sole
non-terminal.
{\color{Myblue}
%\subsection{Generating liveness equations \Lv\  for function bodies}
\paragraph{Handling user-defined functions:}
\label{sec:bodylivenessbodies}
To  avoid analyzing  the body  of a  function for  each
call, we calculated the  liveness for the arguments and
the variables in a function  with respect to a symbolic
demand  $\sigma$.   To  get   the  actual  liveness  we
calculate an  over-approximation of the  actual demands
made by  all the  calls and  calculate the  liveness at
each  GC  point  inside  the  function  based  on  this
approximation.  The {\em summary demand} is
calculated by  taking a union  of the demands  at every
call site of a function analogous to 0-CFA~\cite{Shivers:1988}.

@@

\input{cfg-to-nfa}
}
  \SetStartEndCondition{ }{}{}%
  \SetKwProg{Pro}{procedure}{\string:}{}
  \SetKwProg{Fn}{function}{\string:}{}
  \SetKwFunction{Range}{range}%%
  \SetKw{KwTo}{in}\SetKwFor{For}{for}{\string:}{}%
  \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
  \SetKwFor{While}{while}{:}{fintq}%
  \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
  \SetKwFunction{Lgc}{$\mathsf{lgc}$}%
  \SetKwFunction{Copy}{$\mathsf{copy}$}%
  \SetKwFunction{LCopy}{$lgcCopy$}%
  \SetKwFunction{RCopy}{$rgcCopy$}%
